\textbf{\sc Towards Deliberative Control in Marine Robotics}

### 2012-02-23 trunk/europa.tex
updates on europa.tex for Javier

\eu can ingest descriptions of models, plans and goals and thereby can
be applied to different domains and problems, simply by providing
different models and goals. Consequently the expressiveness of the
language for models, goals and plans is of great importance. For \eue,
\texttt{NDDL} allows the user to specify the different components of a
problem domain in a precise and concise manner. Among the main
features that make \texttt{NDDL} a powerful tool for describing
problem domains are that it is object oriented. \texttt{NDDL} supports
classes and inheritance in similar fashion to popular Object-Oriented
languages like Java and C++. It also supports polymorphism for some of
the planning components. Using object classes and instances is a
time-tested approach to naturally describe problem domains
\cite{rumbaugh}. It offers declarative constructs to define
constraints and causality dependencies between actions and effects and
finally it offers procedural constructs to populate \eus plan
database with specific problem instances expressed in terms of
objects, variables, constraints, facts and goals. To articulate
important elements of \texttt{NDDL}, we show how the Shopping Agent
problem could be modeled:

@@

This problem instance consists of \textbf{Variables} and
\textbf{Objects}, \textbf{Facts}, \textbf{Goals} and
\textbf{Constraints}. In our example we define temporal variables that
are used to specify the planning horizon (\texttt{horizonStart} and
\texttt{horizonEnd}) and objects are used to specify locations,
products, product-location associations and the Shopping Agent that we
want to plan for. Facts are used to specify the current state of the
world and do not need any support and are represented as tokens in
\eus plan database. In this case we use facts to specify that the
Shipping Agent is initially at home. Goals are also represented as
tokens in \eus plan database. Different from facts, goals trigger
inference in \eu which determines whether flaws (unresolved subgoals,
overloaded resources, etc.) need to be resolved; if so, search modules
can be invoked to resolve those flaws as part of the deliberation
process. Finally, constraints are restrictions on and between actions
and goals. In the example above, we use temporal constraints to
specify when facts hold and when the goals should be
accomplished. % As
% always, all constraints in \eus library could be used.

@@

In \eue, all variables and constraints from a problem domain
description are stored inside the \texttt{CRE}.  The \texttt{CRE}
provides a framework where any number of constraint types can
collaborate to perform inference and prune variable domains using
Bounds Propagation and Domain Reduction techniques \kcomment{citation
  needed}. \eu uses a Propagator pattern which keeps track of a set of
constraint instances of the same type and is responsible for ensuring
that whenever any variable is modified or constraints are introduced
or dropped, all related variables are updated through inference.

In a planning problem, variable or constraint changes can come
directly from an end user (by introducing new activities or goals and
relating them to the rest of the plan through constraints, by changing
the start time of an activity, etc), or can come from an automated
planner (by expanding subgoals, introducing precedence constraints to
resolve mutual exclusion conflicts, etc). When these changes happen,
each Propagator is only able to ensure consistency for the constraints
that are associated with it.  To maintain global consistency, the
\texttt{CRE} implements a variant of the \texttt{AC-3} algorithm
\cite{mackworth77} where it asks Propagators to perform inference
until the Constraint Network returns to an stable state.

@@

A \eu user can introduce new constraint types by implementing
specialized Constraint and Propagator classes in C++, however, \eu
provides a built-in Constraint Library with over 25 arithmetic, temporal, set and
resource constraint types so that a large number of problem domains
can be modeled without having to introduce new constraint
types. Tables \ref{tab:calconst}, \ref{tab:setconst},
\ref{tab:objhierarchy} and \ref{tab:misc} show various constraint
types.

Temporal and resource constraints are specially important to model
real-life planning domains, so it is worth going into some more detail
about how \eus built-in support works for those constraint types.

\textit{Temporal Constraints:} \eu uses variables to explicitly
represent timepoints for plan activities and states. Constraints among
timepoints provide a natural way to express domain axioms. For
example, in order to state that activity $A$ must occur before
activity $B$ we can say that the end timepoint of $A$ is $leq$ the
start timepoint of $B$.

\begin{figure}
\centering
\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
\caption{\small Temporal relations defined within the planner are
  based on \texttt{Allen Algebra} relations shown above.}
\label{fig:allen-algebra}
\vskip-0.3cm
\end{figure}

\cite{dechter91} proposed that constraints among timepoints can be
grouped together to form a Simple Temporal Network (STN). Such a
network can be transformed into a Distance Graph (DG) where the
outward arc from a node represents the maximum distance from the
source node to the target node.  The diagram below illustrates a
simple STN with just 2 variables and a single constraint. It also
shows the resulting DG.

reflected by the use of the \texttt{Allen Algebra} \cite{allen84}
temporal relations shown in Fig. \ref{fig:allen-algebra}. Interval
arithmetic in the EUROPA$_2$ domain model using \texttt{Allen Algebra}
allows encoding of temporal constraint relationships between tokens on
concurrent timelines. See \cite{jonsson00} for additional details
including propagation algorithms and interval logic.

@@

\cite{dechter91} also showed that shortest path algorithms could be
used to propagate values in the network and discover a negative
cycle. A negative cycle is a path from a node to itself that has a
path length less than 0. If such a cycle exists, the network is
inconsistent. It was further shown that a single-source shortest path
algorithm was sufficient to detect a negative cycle and provide
sufficient propagation to yield a backtrack-free search. Thus we have
an efficient and complete algorithm for propagating an STN. \eus
Propagator for temporal constraints is a straight implementation of
this algorithm.

\textit{Resource Constraints:} Most real life planning problems have
to deal with the fact that plans made up of actions that require
resources to be performed typically have only a limited amount of
those resources available (think of fuel, storage capacity, etc).  \eu
provides built-in support for resources whose availability can be
measured with real numbers, in terms of \eus representation, resource
object types and their related constraints are provided as follows:

@@

\item \textbf{Reservoir} resources can have both
  \texttt{consume(resource,time,quantity)} and
  \texttt{produce(resource,time,quantity)} constraints. Production and
  consumption is instantaneous (it happens at a single point in time)

\item \textbf{Reusable} resources have
  \texttt{uses(resource,start,duration,quantity)} constraints that use
  a quantity of the resource for a specific duration, then release it.

\item \textbf {Unary} resources are Reusable resources with
  capacity=1, which allows \eu to perform some run time savings
  compared to the generic Reusable class.

@@

\eu offers configuration options so that in addition to the usual
constraint violations when a resource is completely depleted, plan
flaws can be generated whenever a resource level is outside a specific
upper or lower limit, or when instantaneous production or consumption
exceeds a specific threshold.

Recall that \eu supports a flexible temporal representation, where
time is represented by intervals, instead of single values. In this
context, determining whether a resource is the source of a plan flaw
or a constraint violation in general is more difficult than just
computing a profile of the resource and checking against some limits.
In the context of time flexibility, there are many resource profiles
that are possible, depending on when exactly each
production/consumption event happens out of all the possible values
that are in the domain of the variables that represent the time for
the event.  \cite{mus02} proposed an algorithm where the variables
that represent the time for production/consumption events are placed
in a temporal network, similar to the network used by the Temporal
Propagator above, then the constraints on the resource can be seen as
quantities flowing through that network. Computing the maximum flow
will yield the maximum resource usage, and thus a lower bound on the
resource level, and computing a minimum flow will similarly yield an
upper bound. Since maximum flow on a temporal network can be computed
in polynomial time, \eu takes advantage of this algorithm to
efficiently compute upper and lower bound envelopes on the resource
level.

\begin{figure}
\centering
\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
\caption{\small Resource Envelopes computed by  \eu}
\label{fig:resenvelopes}
\vskip+0.1cm
\end{figure}

Resource envelopes can be very useful when trying to figure how
brittle a plan in the face of changing activity durations or start
times. They can also be useful when searching for plans, although they
are relatively expensive to compute so their ability to prune the
search space must be weighted against the ability to search faster
\cite{6007772}.

Given the above elements, an automated problem solver can be
created using:

\item A domain model that describes the variable, object, predicate
  and action types that are relevant for the problem.

\item A problem instance (also called initial state) that consists of:

    \item Variable and Object instances that exist for the entire
      planning horizon.

    \item Temporally scoped predicate and action instances.

    \item Temporally scoped goals.

This information is kept in \eus Plan Database so that inference and
search mechanisms can be used to look for a problem solution.

Given that temporal intervals may extend indefinitely in both positive
and negative directions, for problems with a temporal dimension it is
also common to specify a Planning Horizon when invoking a solver. The
Planning Horizon is the temporal interval that a problem solver has to
worry about, any decisions that fall completely outside of the
Planning Horizon can be ignored. This is very useful for recurrent
activities, consider for instance a model where a medical checkup has
to occur every 3 months, the user can model a recurring activity and
then use the Planning Horizon to ensure that a finite number of
medical checkup activities are generated as part of a plan. This is
also useful to ignore activities that may happen too far in the future
or in the past to be relevant for a particular plan, this way the user
can ensure that the solver is making as few commitments as necessary.

\eu provides a built-in solver that performs Plan Space Planning
\cite{ghallab04} in this approach, the initial state is considered a
partial plan that needs to be refined toward a solution plan that
achieves the goals. The operations to refine the partial plan PP at
any time are:

\item Find the flaws of PP, that is the conditions that prevent it
  from being a solution plan (flaw types are explained below).

\item Select one such flaw

\item Select a resolver for the flaw

\item Refine PP by applying the resolver

\item If all possible resolvers for a particular flaw fail, return
  failure, otherwise continue until resolving all the flaws.

This Plan Space Planning algorithm is translated into \eus
representation as follows:

The initial partial plan is the state of \eus Plan Database after the
initial state has been instantiated, this results in a set of
variable, object and token instances. Inference then takes place to
detect flaws in the partial plan, there are 3 kinds of flaws that can
be detected:

\item Unbound Variable: a variable in the partial plan whose domain is
  not a singleton. Unbound Variables are resolved by specification of
  a value from the domain of the variable.

\item Open Condition: an open condition is an inactive token. Inactive
  tokens can be generated by explicitly posted goals, or when a token
  is activated, rules may fire that create inactive slave tokens as
  described in section \comment{point to EUROPA Plan Representation
    section}.  Open Conditions can be resolved in three ways:

  \item Merging: A free activity is merged with a matching activity
    already in the plan. An activity a is said to match an activity a?
    if a and a? unify and the temporal constraints involving a are
    satisÞed by a?. Thus, a and a? can be considered the same activity
    and we do not need to introduce a in the plan. Consequently, the
    compatibilities associated with a are not Þred, because they have
    been already triggered when a? was introduced in the plan.

  \item Activation: We introduce a new activity a in the current plan
    associating it with the proper timeline, but without choosing a
    speciÞc time slot for it.The compatibilities associated with a are
    applied and the subgoal activities resulting from those
    compatibilities are introduced as free activities. This results in
    both an ordering flaw, corresponding to the just activated
    activity, and a number of open condition flaws, corresponding to
    the new subgoal activities.

  \item Rejection: Some goals may be optional, if that is the case for
    a particular order condition, its corresponding flaw can be
    resolved by discarding the inactive token.

\item Threat: once a token has been placed in the partial plan it may
  impact other tokens indirectly through possible overlapping
  requirements on objects. Recall for example that a token may belong
  to objects (e.g. Timelines) which require a total order over their
  tokens. If any 2 tokens could possibly overlap (though not
  necessarily), then they pose a threat to each other in terms of
  achieving an extension of the current partial plan which is complete
  and consistent. Similarly, threats may arise where tokens share a
  common resource and their current state might yield extensions of
  the current partial plan which are inconsistent. Threats are
  resolved by imposing ordering constraints among tokens.

>>>>

\eu can ingest descriptions of models, plans and goals and thereby can
be applied to different domains and problems, simply by providing
different models and goals. Consequently the expressiveness of the
language for models, goals and plans is of great importance. For \eue,
\texttt{NDDL} allows the user to specify the different components of a
problem domain in a precise and concise manner. Among the main
features that make \texttt{NDDL} a powerful tool for describing
problem domains are that it is object oriented. \texttt{NDDL} supports
classes and inheritance in similar fashion to popular Object-Oriented
languages like Java and C++. It also supports polymorphism for some of
the planning components. Using object classes and instances is a
time-tested approach to naturally describe problem domains
\cite{rumbaugh}. It offers declarative constructs to define
constraints and causality dependencies between actions and effects and
finally it offers procedural constructs to populate \eus \texttt{PDB}
with specific problem instances expressed in terms of objects,
variables, constraints, facts and goals. To articulate important
elements of \texttt{NDDL}, we show how the Shopping Agent problem
could be modeled:

@@

This problem instance consists of \textbf{Variables} and
\textbf{Objects}, \textbf{Facts}, \textbf{Goals} and
\textbf{Constraints}. In our example we define temporal variables that
are used to specify the planning horizon (\texttt{horizonStart} and
\texttt{horizonEnd}) and objects are used to specify locations,
products, product-location associations and the Shopping Agent that we
want to plan for. Facts are used to specify the current state of the
world and do not need any support and are represented as tokens in the
\texttt{PDB}. In this case we use facts to specify that the Shipping
Agent is initially at home. Goals are also represented as tokens in
the \texttt{PDB}. Different from facts, goals trigger inference in \eu
which determines whether flaws (unresolved subgoals, overloaded
resources, etc.) need to be resolved; if so, search modules can be
invoked to resolve those flaws as part of the deliberation
process. Finally, constraints are restrictions on and between actions
and goals. In the example above, we use temporal constraints to
specify when facts hold and when the goals should be
accomplished. % As
% always, all constraints in \eus library could be used.

\begin{figure}[!htb]
\centering
\includegraphics[scale=0.3]{figs/Allen-algebra.pdf}
\caption{\small Temporal relations defined within the planner are
  based on \texttt{Allen Algebra} relations shown above.}
\label{fig:allen-algebra}
\end{figure}

In \eu, problem domains use \texttt{Allen Algebra} \cite{allen84}
temporal relations shown in Fig. \ref{fig:allen-algebra} which allow
encoding of temporal constraint relationships between tokens on
concurrent timelines. \kcomment{This para could do with a little
  embellishment especially since it is the only material that actually
  talks about the kinds of constraints and their origin.}

@@

In \eue, all variables and constraints from a problem domain
description are stored inside the \texttt{CRE}.  The \texttt{CRE}
provides a framework where any number of constraint types can
collaborate to perform inference and prune variable domains using
Bounds Propagation and Domain Reduction techniques \kcomment{citation
  needed}. \eu uses a Propagator pattern which keeps track of a set of
constraint instances of the same type and is responsible for ensuring
variables are inferentially updated with change (modifications or
constraint addition/deletion).

% that whenever any variable is modified or constraints are introduced
% or dropped, all related variables are updated through inference.

In a planning problem, variable or constraint changes can come
directly from an end user (by introducing new activities or goals and
relating them to the rest of the plan through constraints, by changing
the start time of an activity, etc), or from an automated planner (by
expanding subgoals, introducing precedence constraints to resolve
mutual exclusion conflicts, etc). When these changes occur, each
propagator is only able to ensure consistency for the constraints that
are associated with it.  To maintain global consistency, the
\texttt{CRE} implements a variant of the \texttt{AC-3} algorithm
\cite{mackworth77} where propagators are triggered to infer to ensure
the constraint network returns to a quiescent state.

\subsubsection{Constraint Formulation}
\label{sec:europa:constraints}

An \eu user can introduce new constraint types by implementing
specialized constraint and propagator classes; however, \eu provides a
built-in constraint library with a range of arithmetic, temporal, set
and resource constraint types so that a large number of problem
domains can be modeled without having to introduce new constraint
types. Tables \ref{tab:calconst}, \ref{tab:setconst},
\ref{tab:objhierarchy} and \ref{tab:misc} show various constraint
types.

@@

Temporal and resource constraints are especially important to model
real-world planning domains. % so it is worth going into some more
% detail about how \eus built-in support works for those constraint
% types.
\eu uses variables to explicitly represent timepoints for plan
activities and states. Constraints among timepoints provide a natural
way to express domain axioms. For example, in order to state that
activity $A$ must occur before activity $B$ we can say that the end
timepoint of $A$ is $leq$ the start timepoint of $B$.
\cite{dechter91} proposed that constraints among timepoints can be
grouped together to form a Simple Temporal Network
(\texttt{STN}). Such a network can be transformed into a Distance
Graph (\texttt{DG}) where the outward arc from a node represents the
maximum distance from the source node to the target node.
Fig. \ref{fig:stn} illustrates an \texttt{STN} with $2$ variables and
a single constraint along with the resulting \texttt{DG}.
\cite{dechter91} also showed that shortest path algorithms could be
used to propagate values in the network and discover a negative cycle:
namely a path from a node to itself that has a path length less than
$0$. If such a cycle exists, the network is inconsistent. It was
further shown that a single-source shortest path algorithm was
sufficient to detect a negative cycle and provide sufficient
propagation to yield a backtrack-free search. \eus propagator for
temporal constraints is a straight implementation of this algorithm.

@@

Most real-world planning problems need to deal with plans made up of
actions that typically depend upon limited resources (\eg fuel,
storage capacity, battery state of charge, data buffers).  \eu
provides built-in support for resource computation. In \eus
representation resource object types come in three forms with related
constraints. \textbf{Reservoir} resources can have both
\texttt{consume(resource,time,quantity)} and
\texttt{produce(resource,time,quantity)} constraints. Production and
consumption is instantaneous. \textbf{Reusable} resources have
\texttt{uses(resource,start,duration,quantity)} constraints that use a
quantity of the resource for a specific duration, which are then
released. \textbf {Unary} resources are reusable with capacity$ = 1$,
which allows \eu to perform run time savings in comparison to the
generic reusable class. Configuration options are offered such that in
addition to the usual constraint violations when a resource is
completely depleted, plan flaws can be generated whenever a resource
level is outside a specific upper or lower bound, or when
instantaneous production or consumption exceeds a specific threshold.

Since \eu supports a flexible temporal representation, where time is
represented by intervals, instead of single values, determining
whether a resource is the source of a plan flaw or a constraint
violation in general is substantially more difficult than just
computing a profile of the resource and checking against some limits.
With temporal flexibility, a number of resource profiles could be
possible depending on precisely when each production/consumption event
occurs out of a range of possible values that are in the domain of the
variables that represent the temporal extant for the event.
\cite{Muscettola04,Muscettola06} proposed an algorithm where the
variables that represent the time for production/consumption events
are placed in a temporal network, similar to the network used by the
temporal propagator; constraints on the resource can then be seen as
quantities flowing through that network. Computing the maximum flow
then, yields the maximum resource usage and thus a lower bound on the
resource level; computing a minimum flow will similarly yield an upper
bound. Since maximum flow on a temporal network can be computed in
polynomial time, \eu takes advantage of this algorithm to efficiently
compute upper and lower bound envelopes on the resource
level. Fig. \ref{fig:resenvelopes} shows an example of such a resource
computation.

\begin{figure}
\centering
\includegraphics[scale=0.2]{figs/europa-resource-envelopes.png}
\caption{\small An example of resource envelopes computed by \eu.}
\label{fig:resenvelopes}
\end{figure}

Resource envelopes can be useful when trying to evaluate plan
brittleness in the face of changing activity durations or start
times. They can also be useful when searching for plans; however they
are relatively expensive to compute. Therefore their ability to prune
the search space must be weighted against the ability to search faster
\cite{6007772}.

Given the above, an automated problem solver can be created using, a
\emph{domain model} that describes the variable, object, predicate and
action types that are relevant for the problem, a \emph{problem
  instance} or initial state that consists of variable and object
instances that exist for the entire planning horizon, \emph{temporally
  scoped predicate and action} instances and \emph{temporally scoped
  goals}. This information is retained in \eus \texttt{PDB} so that
inference and search mechanisms can be used to look for a problem
solution.

Since temporal intervals may extend indefinitely in both positive and
negative directions, for problems with a temporal dimension it is also
common to specify a \emph{planning horizon} when invoking a solver;
any decisions that fall completely outside of the horizon can be
ignored. This is particularly useful for recurrent activities;
consider for instance a model where a medical checkup has to occur
every $3$ months. The user can model a recurring activity and then use
the horizon to ensure that a finite number of medical checkup
activities are generated as part of a plan. This is also useful to
ignore activities that may occur far in the future or in the past to
be relevant for a particular plan, ensuring that the solver is making
as few commitments as necessary and making planning tractable.

\eu provides a built-in solver that performs \emph{Plan Space
  Planning} \cite{ghallab04}; the initial state is considered a
partial plan that needs to be refined toward a solution that achieves
stated goals. The operations to refine the partial plan \texttt{PP} at
any time are:

@@

\item Find flaws of \texttt{PP}, \ie conditions that prevent it from
  being a solution plan.

\item Select one such flaw.

\item Select a resolver for the flaw.

\item Refine \texttt{PP} by applying the resolver.

\item If all possible resolvers for a particular flaw fail, return
  failure, otherwise continue until resolving all flaws.

@@

% This Plan Space Planning algorithm is translated into \eus
% representation as follows:

The initial partial plan is the state of \eus Plan Database after the
initial state has been instantiated. This results in a set of
variable, object and token instances. Inference then takes place to
detect flaws in the partial plan; the three kinds of flaws that can be
detected are:

\item[\textbf{Unbound Variable}]: a variable in the partial plan whose
  domain is not a singleton. Unbound Variables are resolved by value
  selection from their domain.

\item[\textbf{Open Condition}]: an open condition is an
  \texttt{INACTIVE} token. \texttt{INACTIVE} tokens can be generated
  by explicitly posted goals, or when a token is activated, rules may
  activate the creation of \texttt{INACTIVE} slave tokens as described
  in Section \ref{sec:europa:pr}. Open Conditions can be resolved in
  three ways:

  \item[\textbf{Merging}]: A free activity is merged with a matching
    activity already in the plan. An activity $a$ is said to match an
    activity $a?$ iff $a$ and $a?$ unify \cite{russelnorvig} and the
    temporal constraints involving $a$ are satisfied by $a?$. Thus,
    $a$ and $a?$ can be considered the same activity without the need
    to introduce $a$ in the plan. Consequently, the compatibilities
    \kcomment{compatibilities have not been defined} associated with
    $a$ are not Þred \kcomment{???}, because they have been already
    triggered when $a?$ was introduced in the plan.

  \item[\textbf{Activation}]: A new activity $a$ is introduced in the
    current plan associating it with the correct timeline, without
    choosing a specific time slot. The compatibilities \kcomment{as
      above} associated with $a$ are applied and the subgoal
    activities resulting from those compatibilities \kcomment{as
      above} are introduced as free activities \kcomment{what is a
      free activity?}. This results in both an ordering flaw
    \kcomment{undefined}, corresponding to the just activated
    activity, and a number of open condition flaws, corresponding to
    the new subgoal activities.

  \item[\textbf{Rejection}]: Some goals may be optional; if so its
    corresponding flaw can be resolved by discarding the
    \texttt{INACTIVE} token.

\item[\textbf{Threat}]: Once a token has been placed in the partial
  plan it may impact other tokens indirectly through possible
  overlapping requirements on objects. Recall for example that a token
  may belong to objects (\eg timelines) which require a total order
  over their tokens. If any two tokens could possibly overlap (though
  not necessarily), then they pose a threat to each other in terms of
  achieving an extension of the current partial plan which is complete
  and consistent. Similarly, threats may arise where tokens share a
  common resource and their current state might yield extensions of
  the current partial plan which are inconsistent. Threats are
  resolved by imposing ordering constraints among tokens.
